Forecasting the 2024 US Presidential Election

### 2024-10-24 paper/paper.qmd
Updated data section

# Data {#sec-data}

## Overview
We sourced the "Presidential General Election Polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and performed an in-depth analysis using the statistical programming language R [@citeR].

@@

# Name references
column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  young_age_18_29 = "What is your age range? 18-29 years",
  young_age_30_39 = "What is your age range? 30-39 years",
  young_age_40_49 = "What is your age range? 40-49 years",
  older_age_50_59 = "What is your age range? 50-59 years",
  older_age_60_69 = "What is your age range? 60-69 years",
  older_age_70_more = "What is your age range? 70 or more years"
)

@@

poll_summary <- poll_data %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct,
    age_young_pct = mean(`What is your age range? 18-29 years`, na.rm = TRUE) + 
                    mean(`What is your age range? 30-39 years`, na.rm = TRUE) + 
                    mean(`What is your age range? 40-49 years`, na.rm = TRUE),
    older_pct = mean(`What is your age range? 50-59 years`, na.rm = TRUE) + 
                mean(`What is your age range? 60-69 years`, na.rm = TRUE) + 
                mean(`What is your age range? 70 or more years`, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to nearest tenth

colnames(poll_summary) <- c("State", "Democrat Pct", "Republican Pct", 
                            "High Educated", "Low Educated", "Female", 
                            "Male", "Nonbinary", "Caucasian", 
                            "Minority Ethnicity", "Young Age", "Older Age")
poll_summary_party <- poll_summary %>%
  select(State, `Democrat Pct`, `Republican Pct`)

poll_summary_demographics <- poll_summary %>%
  select(State, `High Educated`, `Low Educated`, `Female`, 
         `Male`, `Nonbinary`, `Caucasian`, 
         `Minority Ethnicity`, `Young Age`, `Older Age`)

kable(poll_summary_party) %>%
  kable_styling(font_size = 10, position = "center", full_width = FALSE)

@@

```
\newpage  
## Measurement
For this analysis, we selected Emerson as our primary pollster. The rationale behind this choice is explained in detail in @sec-appenA. Emerson’s polls provided essential details, including polling date ranges, party affiliations, sample sizes, and the percentage breakdown of support for different political parties.

@@

By visualizing the difference between Democratic and Republican preferences across states, we are able to pinpoint which states are firmly partisan and which fall closer to the center, indicating their potential to swing either way in future elections.

@@

Next, we take a deeper dive into the various demographic and social factors that influence voter preferences. The predictors we consider include education levels, gender, ethnicity, and age group. These variables are crucial as they shape political preferences and are likely to influence which way a state leans politically.

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    non_binary_other_pct = round(mean(.data[[column_refs$non_binary_other]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1),
    age_young_pct = round(mean(.data[[column_refs$young_age_18_29]], na.rm = TRUE) + 
                          mean(.data[[column_refs$young_age_30_39]], na.rm = TRUE) + 
                          mean(.data[[column_refs$young_age_40_49]], na.rm = TRUE), 1),
    older_pct = round(mean(.data[[column_refs$older_age_50_59]], na.rm = TRUE) + 
                      mean(.data[[column_refs$older_age_60_69]], na.rm = TRUE) + 
                      mean(.data[[column_refs$older_age_70_more]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    NonBinary = non_binary_other_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct,
    YoungAge = age_young_pct,
    OlderAge = older_pct
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Predictors and Party Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1),  # Axis labels size
  )

>>>>

\newpage  
# Data {#sec-data}

## Overview

@@

@@

poll_summary <- poll_data %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to nearest tenth

colnames(poll_summary) <- c("State", "Democrat Pct", "Republican Pct", 
                            "High Educated", "Low Educated", "Female", 
                            "Male", "Nonbinary", "Caucasian", 
                            "Minority Ethnicity")

poll_summary_party <- poll_summary %>%
  select(State, `Democrat Pct`, `Republican Pct`)

poll_summary_demographics <- poll_summary %>%
  select(State, `High Educated`, `Low Educated`, `Female`, 
         `Male`, `Nonbinary`, `Caucasian`, 
         `Minority Ethnicity`)

# Table for party percentages
kable(poll_summary_party) %>%
  kable_styling(font_size = 10, position = "center", full_width = FALSE)

@@

```
We sourced the "Presidential General Election Polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and performed an in-depth analysis using the statistical programming language R [@citeR].

## Measurement  
For this analysis, we selected Emerson as our primary pollster. The rationale behind this choice is explained in detail in @sec-appenA. Emerson’s polls provided essential details, including polling date ranges, party affiliations, sample sizes, and the percentage breakdown of support for different political parties.

@@

```
```{r}
#| label: fig-party-extremes
#| fig-cap: Demographic Differences Between Top Red and Blue States
#| echo: false
#| warning: false
#| message: false

library(tidyverse)

# Prepare the poll_summary data frame
poll_summary <- poll_data %>%
  mutate(party_diff = democrat_pct - republican_pct) %>%
  select(-c(2, 3, 4, 6, 7, 8, 9, 11)) %>%  # Remove specific columns by index
  select(-contains('Mode')) %>%             # Remove columns containing 'Mode' in their names
  select(-contains('registration')) %>%     # Remove columns containing 'registration' in their names
  drop_na()                                 # Remove rows with any NA values

# Function to identify top N blue and N red states
identify_top_states <- function(data, n) {
  data %>%
    mutate(party_diff = democrat_pct - republican_pct) %>%
    arrange(desc(party_diff)) %>%
    summarise(
      blue_states = list(state[1:n]),  # Use 'n' instead of hardcoding 4
      red_states = list(state[(nrow(data) - n + 1):nrow(data)])
    )
}

# Function to compute mean differences for all numeric variables
compute_mean_differences <- function(data, blue_states, red_states) {
  # Select numeric variables excluding 'state' and 'party_diff'
  numeric_vars <- data %>%
    select(-state, -party_diff) %>%
    select(where(is.numeric)) %>%
    names()

  # Compute means for blue states
  blue_means <- data %>%
    filter(state %in% blue_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  # Compute means for red states
  red_means <- data %>%
    filter(state %in% red_states) %>%
    summarise(across(all_of(numeric_vars), mean, na.rm = TRUE))

  # Compute absolute differences
  diff_df <- tibble(
    Variable = numeric_vars,
    Blue_Mean = as.numeric(blue_means[1, ]),
    Red_Mean = as.numeric(red_means[1, ]),
    Abs_Difference = abs(as.numeric(blue_means[1, ]) - as.numeric(red_means[1, ]))
  )

  # Remove variables where both means are NA
  diff_df <- diff_df %>%
    filter(!is.na(Blue_Mean) & !is.na(Red_Mean))

  diff_df
}

# Function to prepare data for plotting
prepare_plot_data <- function(diff_df, top_n = 6) {
  # Get top N variables with the largest differences
  top_diff_vars <- diff_df %>%
    arrange(desc(Abs_Difference)) %>%
    slice(1:top_n)

  # Prepare data for plotting
  plot_data <- top_diff_vars %>%
    select(Variable, Blue_Mean, Red_Mean) %>%
    pivot_longer(
      cols = c("Blue_Mean", "Red_Mean"), 
      names_to = "Group", 
      values_to = "Mean"
    ) %>%
    mutate(Group = recode(Group, "Blue_Mean" = "Blue States", "Red_Mean" = "Red States"))

  plot_data
}

# Function to create the plot with y-axis from 0 to 100 and data labels
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    scale_fill_manual(
      values = c("Blue States" = "red", "Red States" = "blue")  # Reversed colors
    ) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

# Main analysis function
analyze_state_differences <- function(data, top_n_states = 1, top_n_vars = 6) {
  top_states <- identify_top_states(data, n = top_n_states)

  diff_df <- compute_mean_differences(
    data, 
    blue_states = top_states$blue_states[[1]], 
    red_states = top_states$red_states[[1]]
  )

  plot_data <- prepare_plot_data(diff_df, top_n = top_n_vars)

  plot <- create_comparison_plot(plot_data)

  list(
    plot = plot,
    top_variables = diff_df %>% arrange(desc(Abs_Difference)) %>% slice(1:top_n_vars)
  )
}

# Ensure 'state' column exists in poll_summary
if (!"state" %in% colnames(poll_summary)) {
  stop("The 'state' column does not exist in poll_summary.")
}

# Run the analysis
analysis_results <- analyze_state_differences(poll_summary, top_n_states = 1, top_n_vars = 6)

# Display the plot
analysis_results$plot

```
Beyond identifying swing states, it's also important to understand the differences in demographics between the most strongly Democratic ("blue") and Republican ("red") states. The bar graph in @fig-party-extremes takes a closer look at these differences. It focuses on key factors that show the biggest contrasts between Democratic-leaning and Republican-leaning states, helping to highlight how these factors differ between the most extreme red and blue states.

The @fig-party-extremes visualization provides insight into how people's responses to polls may influence their political preferences. This extra layer of analysis helps clarify the factors shaping the political leanings of different states and adds depth to our understanding of the electorate. In this case, we examine people's opinions on the current president and their voting choices in past elections.

Through these visualizations, we not only identify which states are most likely to swing in future elections but also gain a deeper understanding of the public's opinions that contribute to these partisan divides.

@@

Next, we take a deeper dive into the various demographic and social factors that influence voter preferences. The predictors we consider include education levels, gender, ethnicity, and age group. These variables are crucial as they shape political preferences and are likely to influence which way a state leans politically.  
```{r}
#| label: updating-predictors
#| echo: false

column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  hispanic_latino = "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  vote_trump = "Who did you vote for in the 2020 election? Donald Trump",
  vote_biden = "Who did you vote for in the 2020 election? Joe Biden",
  biden_approve = "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  biden_disapprove = "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove",
  young_age_18_29 = "What is your age range? 18-29 years",
  young_age_30_39 = "What is your age range? 30-39 years",
  young_age_40_49 = "What is your age range? 40-49 years",
  older_age_50_59 = "What is your age range? 50-59 years",
  older_age_60_69 = "What is your age range? 60-69 years",
  older_age_70_more = "What is your age range? 70 or more years"
)
```

@@

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    non_binary_other_pct = round(mean(.data[[column_refs$non_binary_other]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    NonBinary = non_binary_other_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct
    # No age-related columns included anymore
  )

@@

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Predictors and Party Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )
```

```{r}
#| label: fig-predictors-party-pct-2
#| fig-cap: Example Voting and Approval Predictor Variable Visualization
#| echo: false

# Calculate poll_summary with rounding to the nearest tenth
poll_summary_2 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1),
    biden_approve_pct = round(mean(.data[[column_refs$biden_approve]], na.rm = TRUE), 1),
    biden_disapprove_pct = round(mean(.data[[column_refs$biden_disapprove]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data_2 <- poll_summary_2 %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    VotedTrump2020 = vote_trump_pct,
    VotedBiden2020 = vote_biden_pct,
    ApprovesOfBiden = biden_approve_pct,
    DisapprovesOfBiden = biden_disapprove_pct
  )

# Calculate the correlation matrix
cor_matrix_2 <- cor(corr_data_2, use = "complete.obs")

# Set correlations less than -0.9 to NA
cor_matrix_2[cor_matrix_2 < -0.9] <- NA

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix_2, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Voting Preferences and State Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )

```{r}
#| label: fig-predictors-education-race-voting
#| fig-cap: Correlation Matrix of Education, Race, and Past Voting Behavior
#| echo: false

# Calculate poll_summary with rounding to the nearest tenth
poll_summary_3 <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1),
    vote_trump_pct = round(mean(.data[[column_refs$vote_trump]], na.rm = TRUE), 1),
    vote_biden_pct = round(mean(.data[[column_refs$vote_biden]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data_3 <- poll_summary_3 %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct,
    VotedTrump2020 = vote_trump_pct,
    VotedBiden2020 = vote_biden_pct
  )

# Calculate the correlation matrix
cor_matrix_3 <- cor(corr_data_3, use = "complete.obs")

# Set correlations less than -0.9 to NA
cor_matrix_3[cor_matrix_3 < -0.85] <- NA

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix_3, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Education, Race, and Past Voting Behavior",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1)  # Axis labels size
  )
